---
title: "Application and available datasets for \n the analysis of distributional SD"
author: "A. Irpino, R. Verde<br>ESTP Cologne 14-16 May 2024"
format: 
  revealjs: 
    theme: [dark, custom.scss]
    width: 1280
    height: 720
    fontsize: 2em
    linestretch: 1.1
    footer: "Applications and datasets for distributional SD, ESTP Cologne 14-16 May 2024"
    slideNumber: true
filters:
  - "parse-latex.lua"
title-slide-attributes:
  data-background-image: "images/main.svg"
  data-background-color: white
  data-background-size: contain
bibliography: biblio.bib
link-citations: true
editor: visual
---

```{r}
source("../code/useful_funs.R")
```

# Datasets available for exercitating

## Datasets contained in the `HistDAWass` package

The `HistDAWass` package in R contains some distributional datasets generated from some books and from raw data from different contexts of application. Let's see some of them:

- `BLOOD` dataset, is the description of 14 typologies of patients in a hospital, the variables are histograms of the levels o Hematocrit, Hemoglobin and Cholesterol
- `Age_Pyramids_2014`	Age pyramids of all the countries of the World in 2014
- `China_Month`	Distributions of climatic variables for each month  of 60 stations
- `China_Seas`	Distributions of climatic variables for each season  of 60 stations China
- `OzoneFull`	Full Ozone dataset for Histogram data analysis

## The BLOOD dataset

```{r fig.align='center'}
library(HistDAWass)
plot(BLOOD)
```

-----------

### The AGE PYRAMIDS dataset (229 counties)

::: columns
::: {.column width="25%"}

```{r, fig.dim=c(8,10),out.width = "100%"}
plot(Age_Pyramids_2014[1:60,], type="DENS")
```

:::
::: {.column width="25%"}
```{r, fig.dim=c(8,10),out.width = "100%"}
plot(Age_Pyramids_2014[61:120,], type="DENS")
```
:::
::: {.column width="25%"}
```{r, fig.dim=c(8,10),out.width = "100%"}
plot(Age_Pyramids_2014[121:180,], type="DENS")
```
:::
::: {.column width="25%"}
```{r, fig.dim=c(8,10),out.width = "100%"}
plot(Age_Pyramids_2014[181:229,], type="DENS")
```
:::
:::

---------

### The CHINA season dataset 60 obs, 56 vars

```{r cache=T}
p1<-plot(China_Seas[,1:28], type="DENS")
p2<-plot(China_Seas[,29:56], type="DENS")
```


```{r, fig.dim=c(24,8),out.width="100%",cache=T}
p1
```

-----------------

```{r, fig.dim=c(24,8),out.width="100%",cache=T}
p2
```

---------

### The OZONE dataset 78 obs, 4 vars

::: columns
::: {.column width="50%"}

```{r, fig.dim=c(8,8),out.width = "80%", fig.align='center'}
plot(OzoneFull[1:39,], type="DENS")
```

:::
::: {.column width="50%"}
```{r, fig.dim=c(8,8),out.width = "80%", fig.align='center'}
plot(OzoneFull[39:78,], type="DENS")
```
:::
:::

----------------

#### Other distributional datasets {.smaller}

- `HMAT` Time use distributional data created from from an IPUMS dataset available from <https://github.com/Airpino/Symbolic_data_analysis_softw/raw/main/Histo_data_IPUMS.RData>

```{r, fig.dim=c(30,10),out.width = "100%", fig.align='center'}
load("../Histo_data_IPUMS.RData")
plot(HMAT, type="DENS")
```


# Some exploratory tools

## Some links showing some exploration tools for distributional data

- An application on climatic data: <https://airpino.github.io/ICDS_23_presentation/#/title-slide>

- An application on EUSILC 2013 data: <https://airpino.github.io/DSSR2024_pres/#/title-slide>


## An analysis task

1. Choose a dataset 
2. Explore the data
    - compute some basic statistics from data
3. Fix some objectives
4. Analyze the data (or a selection of the data) using at least one of the following techniques
   - Principal components analysis
   - Regression analysis
   - Clustering analysis
5. Present the results
6. Let's discuss together